dynamics

### 2023-02-15 scripts/workshop/dynamics.Rmd
knit it

```{r}
library(tidyverse)
library(magrittr)
#install_github("thomasp85/patchwork")
library(patchwork)
library(ggnewscale)
library(here)
```

@@

halflife <- contour(trans_1_min = 0.0001,trans_1_max = 0.001,trans_3_min = 0.0001,trans_3_max = 0.001,mort1 = c(1/40),mort2 = 1/40,degr = 1/3.895598)

@@

>>>>

```{r}
library(tidyverse)
library(magrittr)
#install_github("thomasp85/patchwork")
library(patchwork)
library(ggnewscale)
library(here)
library(deSolve)
```

@@

halflife <- contour(trans_1_min = 0.0001,trans_1_max = 0.001,trans_3_min = 0.0001,trans_3_max = 0.001,mort1 = c(1/45),mort2 = 1/45,degr = 1/3.895598)

@@

```{r}
rm0<-function(t,x,params){
  Sa <- x[1]  
  Ia <- x[2]
  Ra <- x[3]
  Sb <- x[4]  
  Ib <- x[5]
  Rb <- x[6]
  V <- x[7]
  with(as.list(params),{
    dSa <- -beta*Sa*Ia - betab*Sa*Ib - phi*Sa*V + lambda - mu_a*Sa
    dIa <- beta*Sa*Ia + betab*Sa*Ib + phi*Sa*V - alpha*Ia - mu_a*Ia
    dRa <- alpha*Ia - mu_a*Ra

    dSb <- -betab*Sb*Ib - beta*Sb*Ia - phi*Sb*V + lambda - mu_b*Sb
    dIb <- betab*Sb*Ib + beta*Sb*Ia + phi*Sb*V - alpha*Ib - mu_b*Ib
    dRb <- alpha*Ib - mu_b*Rb

    dV <- sigma*Ia + sigma*Ib - epsilon*V
    res<-c(dSa,dIa,dRa,dSb,dIb,dRb,dV)
    list(res)
  })}
maxTime <- 500.0 # time is in years - run model for this time
times<-seq(0,maxTime,by=1) # how long we run the model for
# notes on params
# mu =     <- density-independent mortality rate
# beta =       <- contact transmission rate by infected A
# betab =       <- contact transmission rate by infected B
# phi =      <- environmental transmission rate
# lambda =       <- density-dependent growth rate
# alpha =     <- recovery rate
# sigma =    <- shedding rate
# epsilon =       <- viral degradation rate
params<-c(mu_a=1/50,
          mu_b=1/50,
          beta=0.001,
          betab=0.0001,
          phi=0.001,
          lambda=5/3,
          alpha=1/10,
          sigma=1/2,
          epsilon=1/2)  # model parameters

xstart<-c(Sa=116,
          Ia=1,
          Ra=0,
          Sb=83,
          Ib=1,
          Rb=0,
          V=0)  # initial conditions

output<-as.data.frame(lsoda(xstart,times,rm0,params)) # tells computer to solve (integrate) equations

output %>% ggplot(.,aes(x=time))+
  geom_line(aes(y=Sa,col="Susceptible A"))+
  geom_line(aes(y=Ia,col="Infected A"))+
  geom_line(aes(y=Ra,col="Recovered A"))+
  geom_line(aes(y=Sb,col="Susceptible B"))+
  geom_line(aes(y=Ib,col="Infected B"))+
  geom_line(aes(y=Rb,col="Recovered B"))+
  geom_line(aes(y=V,col="Virus"))+
  scale_colour_manual(values = c("red","orange","yellow","green","blue","black","violet"))+
  theme(legend.position = c(0.85,0.5))+
  labs(y="N",x="Time",col="Population")+
  ggtitle("2-Species Transmission Model for Ranavirus in an Amphibian Wetland")+theme(plot.title = element_text(hjust=0, size=10))
```

notes on params
mu =     <- density-independent mortality rate
beta =       <- contact transmission rate by infected A
betab =       <- contact transmission rate by infected B
phi =      <- environmental transmission rate
lambda =       <- density-dependent growth rate
alpha =     <- recovery rate
sigma =    <- shedding rate
epsilon =       <- viral degradation rate

transmission parameter values
species 1 is held constant at 0.0001
second species is more competent at 0.0004
env. trans rate is 0.00058

```{r}
spec_1 <- 0.0001
spec_2 <- 0.00055
env_rate <- 0.00065
```

Neutral
```{r}
params_n <- c(mu_a=1/50,
          mu_b=1/50,
          beta=spec_1,
          betab=spec_2,
          phi=env_rate,
          lambda=5/3,
          alpha=1/10,
          sigma=1/2,
          epsilon=1/1.947799)  # model parameters

xstart_n <- c(Sa=params_n[[6]]/params_n[[1]],
          Ia=1,
          Ra=0,
          Sb=params_n[[6]]/params_n[[2]],
          Ib=1,
          Rb=0,
          V=0)  # initial conditions

output_n <- as.data.frame(lsoda(xstart_n,times,rm0,params_n)) # tells computer to solve (integrate) equations
```

```{r}
output_n %>% ggplot(.,aes(x=time))+
  geom_line(aes(y=Sa,col="Susceptible A"))+
  geom_line(aes(y=Ia,col="Infected A"))+
  geom_line(aes(y=Ra,col="Recovered A"))+
  geom_line(aes(y=Sb,col="Susceptible B"))+
  geom_line(aes(y=Ib,col="Infected B"))+
  geom_line(aes(y=Rb,col="Recovered B"))+
  geom_line(aes(y=V,col="Virus"))+
  scale_colour_manual(values = c("red","orange","yellow","green","blue","black","violet"))+
  theme(legend.position = c(0.85,0.5))+
  labs(y="N",x="Time",col="Population")+
  ggtitle("2-Species Transmission Model for Ranavirus in an Amphibian Wetland - Neutral")+theme(plot.title = element_text(hjust=0, size=10))
```

Community Composition
```{r}
params_c <- c(mu_a=1/60,
          mu_b=1/30,
          beta=spec_1,
          betab=spec_2,
          phi=env_rate,
          lambda=5/3,
          alpha=1/10,
          sigma=1/2,
          epsilon=1/1.947799)  # model parameters

output_c <- as.data.frame(lsoda(xstart_n,times,rm0,params_c)) # tells computer to solve (integrate) equations

```{r}
output_c %>% ggplot(.,aes(x=time))+
  geom_line(aes(y=Sa,col="Susceptible A"))+
  geom_line(aes(y=Ia,col="Infected A"))+
  geom_line(aes(y=Ra,col="Recovered A"))+
  geom_line(aes(y=Sb,col="Susceptible B"))+
  geom_line(aes(y=Ib,col="Infected B"))+
  geom_line(aes(y=Rb,col="Recovered B"))+
  geom_line(aes(y=V,col="Virus"))+
  scale_colour_manual(values = c("red","orange","yellow","green","blue","black","violet"))+
  theme(legend.position = c(0.85,0.5))+
  labs(y="N",x="Time",col="Population")+
  ggtitle("2-Species Transmission Model for Ranavirus in an Amphibian Wetland - Community Composition")+theme(plot.title = element_text(hjust=0, size=10))
```

Abundance
```{r}
params_a <- c(mu_a=1/52.5,
          mu_b=1/52.5,
          beta=spec_1,
          betab=spec_2,
          phi=env_rate,
          lambda=5/3,
          alpha=1/10,
          sigma=1/2,
          epsilon=1/1.947799)  # model parameters

output_a <- as.data.frame(lsoda(xstart_n,times,rm0,params_a)) # tells computer to solve (integrate) equations
```

```{r}
output_a %>% ggplot(.,aes(x=time))+
  geom_line(aes(y=Sa,col="Susceptible A"))+
  geom_line(aes(y=Ia,col="Infected A"))+
  geom_line(aes(y=Ra,col="Recovered A"))+
  geom_line(aes(y=Sb,col="Susceptible B"))+
  geom_line(aes(y=Ib,col="Infected B"))+
  geom_line(aes(y=Rb,col="Recovered B"))+
  geom_line(aes(y=V,col="Virus"))+
  scale_colour_manual(values = c("red","orange","yellow","green","blue","black","violet"))+
  theme(legend.position = c(0.85,0.5))+
  labs(y="N",x="Time",col="Population")+
  ggtitle("2-Species Transmission Model for Ranavirus in an Amphibian Wetland - Abundance")+theme(plot.title = element_text(hjust=0, size=10))
```

Half-life
```{r}
params_h <- c(mu_a=1/45,
          mu_b=1/45,
          beta=spec_1,
          betab=spec_2,
          phi=env_rate,
          lambda=5/3,
          alpha=1/10,
          sigma=1/2,
          epsilon=1/3.895598)  # model parameters

output_h <- as.data.frame(lsoda(xstart_n,times,rm0,params_h)) # tells computer to solve (integrate) equations
```

```{r}
output_h %>% ggplot(.,aes(x=time))+
  geom_line(aes(y=Sa,col="Susceptible A"))+
  geom_line(aes(y=Ia,col="Infected A"))+
  geom_line(aes(y=Ra,col="Recovered A"))+
  geom_line(aes(y=Sb,col="Susceptible B"))+
  geom_line(aes(y=Ib,col="Infected B"))+
  geom_line(aes(y=Rb,col="Recovered B"))+
  geom_line(aes(y=V,col="Virus"))+
  scale_colour_manual(values = c("red","orange","yellow","green","blue","black","violet"))+
  theme(legend.position = c(0.85,0.5))+
  labs(y="N",x="Time",col="Population")+
  ggtitle("2-Species Transmission Model for Ranavirus in an Amphibian Wetland - Half-life")+theme(plot.title = element_text(hjust=0, size=10))
```

Combined
```{r}
params_x <- c(mu_a=1/70,
          mu_b=1/35,
          beta=spec_1,
          betab=spec_2,
          phi=env_rate,
          lambda=5/3,
          alpha=1/10,
          sigma=1/2,
          epsilon=1/3.895598)  # model parameters

output_x <- as.data.frame(lsoda(xstart_n,times,rm0,params_x)) # tells computer to solve (integrate) equations
```

```{r}
output_x %>% ggplot(.,aes(x=time))+
  geom_line(aes(y=Sa,col="Susceptible A"))+
  geom_line(aes(y=Ia,col="Infected A"))+
  geom_line(aes(y=Ra,col="Recovered A"))+
  geom_line(aes(y=Sb,col="Susceptible B"))+
  geom_line(aes(y=Ib,col="Infected B"))+
  geom_line(aes(y=Rb,col="Recovered B"))+
  geom_line(aes(y=V,col="Virus"))+
  scale_colour_manual(values = c("red","orange","yellow","green","blue","black","violet"))+
  theme(legend.position = c(0.85,0.5))+
  labs(y="N",x="Time",col="Population")+
  ggtitle("2-Species Transmission Model for Ranavirus in an Amphibian Wetland - Combined")+theme(plot.title = element_text(hjust=0, size=10))
```
